---
title: 'Data Science Project 1'
author: 'YOUR NAMES HERE'
date: 'Updated: `r format(Sys.time(), "%A, %B %d, %Y @ %X")`'
output: html_document
---

<hr> 

#### DATA FILES

* [Mauna Loa Monthly CO2 dataset](https://raw.githubusercontent.com/STAT-JET-ASU/DataScience1/master/Projects/CO2monthlyM.csv)
* [Mauna Loa Weekly CO2 dataset](https://raw.githubusercontent.com/STAT-JET-ASU/DataScience1/master/Projects/CO2weeklyM.csv)
* [Global Monthly CO2 dataset](https://raw.githubusercontent.com/STAT-JET-ASU/DataScience1/master/Projects/CO2monthlyG.csv)
* [Global Daily CO2 dataset](https://raw.githubusercontent.com/STAT-JET-ASU/DataScience1/master/Projects/CO2dailyG.csv)

Refer to the Mauna Loa data [source files](https://www.esrl.noaa.gov/gmd/ccgg/trends/data.html) and global data [source files](https://www.esrl.noaa.gov/gmd/ccgg/trends/gl_data.html) for more information on how the data were collected and descriptions of the variables included in each dataset.

```{r}
# Read in the datasets and load any necessary packages here
# Suppress the code and any messages / warnings it produces

```


#### DATA EXPLORATIONS

* Write a brief description of each dataset's origin and variables, similar to  [here](https://stat-jet-asu.github.io/Datasets/InstructorDescriptions/galtonheightdata.html) and [here](https://stat-jet-asu.github.io/Datasets/InstructorDescriptions/unemployment.html).

* Run `glimpse()` and `summary()` on each dataset to examine its structure and contents.

##### Mauna Loa Monthly

```{r}

```

##### Mauna Loa Weekly

```{r}

```

##### Global Monthly

```{r}

```

##### Global Daily

```{r}

```


#### DATA VISUALIZATIONS

* Replicate the plots shown on [this web page](https://www.esrl.noaa.gov/gmd/ccgg/trends/full.html) and [this web page](https://www.esrl.noaa.gov/gmd/ccgg/trends/index.html). You do not need to include the NOAA / Scripps logos or labels. Your results should look similar to Dr. Thomley's replications [here](https://stat-jet-asu.github.io/DataScience1/Projects/Project01ThomleyPlots.html).

* Create time series plots to show the "full record" and "last five years" for the global monthly CO~2~ data. Make your own choices with regard to axis formatting, line and point styles, colors, labels, etc. 

* Use the weekly Mauna Loa data to create a plot showing the change in CO2 for one year, 10 years, and since 1800 for the years 2000 to present. You will need to create new variables for the 1-year and 10-year change in CO~2~.

* Use the `grid.arrange()` function from the `gridExtra` package to create a display that includes the following three plots stacked on top of one another. Exclude the incomplete 2018 data from all plots.

    + Using daily global data, create side-by-side boxplots of CO~2~ by year. Include a horizontal line at 400ppm, which is considered by many to be a symbolic threshold CO~2~ value for global warming/climate change.
    + Using daily global data, create a barplot showing the mean CO~2~ for each year. Include a horizontal line at 280ppm (approximate pre-industrial CO~2~ average) and at 200ppm (approximate ice age CO~2~ average).
    + Using the global growth data, create a barplot of growth rates for the same time period shown in the other two plots, including error bars to indicate the degree of uncertainty in the estimates.

```{r}
# Put the code for each bullet in a separate code chunk
# Do not suppress your code here when you knit to HTML

```


#### QUESTIONS

1) What trend(s) or patterns do you observe with regard to CO~2~ concentration over time?

ANSWER: 

2) In what way could these analyses be used to support the theory of anthropogenic (man-made) climate change? 

ANSWER: 

3) Why are data and graphs such as these _evidence_ rather than _proof_ of anthropogenic climate change?

ANSWER: 

<hr>

DONE!!!

