---
title: 'Data Science Project 1'
author: 'YOUR NAMES HERE'
date: 'Updated: `r format(Sys.time(), "%A, %B %d, %Y @ %X")`'
output: html_document
---

<hr> 

#### DATA FILES

* [Source and description of MLCO2annual dataset](ftp://aftp.cmdl.noaa.gov/products/trends/co2/co2_annmean_mlo.txt)

* [MLC02annual dataset](https://raw.githubusercontent.com/STAT-JET-ASU/DataScience1/master/Projects/MLCO2annual.csv) (comma delimited text file)

* [Source and description of MLCO2monthly dataset](ftp://aftp.cmdl.noaa.gov/products/trends/co2/co2_mm_mlo.txt)

* [MLCO2monthly dataset](https://raw.githubusercontent.com/STAT-JET-ASU/DataScience1/master/Projects/MLCO2monthly.csv) (comma delimited text file)

* [Source and description of MLCO2weekly dataset](ftp://aftp.cmdl.noaa.gov/products/trends/co2/co2_weekly_mlo.txt)

* [MLCO2weekly dataset](https://raw.githubusercontent.com/STAT-JET-ASU/DataScience1/master/Projects/MLCO2weekly.csv) (comma delimited text file)


```{r}
# Read in the datasets and load any necessary packages here
# Suppress the code and any messages / warnings it produces

```

#### PRODUCTS

Use only the three datasets provided. In some cases, you will have to create new variables or perform your own calculations / transformations. When replicating graphs, you do not have to include the round NOAA and Scripps badges or the March 2018 date stamps.

* Replicate the content of the graph shown on [this page](https://www.esrl.noaa.gov/gmd/ccgg/trends/index.html).

* Replicate the content of the graph shown on [this page](https://www.esrl.noaa.gov/gmd/ccgg/trends/full.html).

* Replicate the content of the graph shown on [this page](https://www.esrl.noaa.gov/gmd/ccgg/trends/gr.html).

* Using the monthly data, create side-by-side boxplots of CO2 by decade. Exclude the 1950s (1959 and before) and 2010s (2010 and after) because the data for those decades is incomplete.

* Using the weekly data, use the [`multiplot`](http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/) function from the `Rmisc` package to create a display for the 21st century that includes three stacked plots. Exclude the incomplete 2018 data.
    + side-by-side boxplots of CO2 by year
    + side-by-side boxplots of adjusted CO2 by year
    + a time series plot showing the change in CO2 for one year, 10 years, and since 1800

```{r}
# Put the code for each bullet in a separate code chunk
# Do not suppress your code here when you knit to HTML

```

#### QUESTION

In what way could these analyses be used to support the theory of anthropogenic (man-made) climate change? Why are data and graphs such as these *evidence* rather than *proof*?


<hr>

DONE!!!

