---
title: "Making Maps in R"
author: "Jill Thomley"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(comment = "", message = FALSE, warning = FALSE)
```

## Mapping in R

```{r}
library(tidyverse)
library(maps)
```

Let's go into the `maps` package and see what's available!

# Some Mapping Basics

## Dataset `usa`

```{r}
usa <- map_data("usa")
head(usa, 10)
```

## Structure

```{r}
str(usa)
table(usa$group)
```

## Mapping `usa`

```{r, fig.height = 3.5, fig.width = 6, fig.align = "center"}
ggplot(usa) + 
  geom_polygon(aes(x = long, y = lat, group = group), 
               color = "darkgray", fill = "lightgray")
```

## A Different Projection

```{r, fig.height = 3.5, fig.width = 6, fig.align = "center"}
ggplot(usa) + 
  geom_polygon(aes(x = long, y = lat, group = group), 
               color = "darkgray", fill = "lightgray") +
  coord_map()
```

## Another Projection

```{r, fig.height = 3.5, fig.width = 6, fig.align = "center"}
ggplot(usa) + 
  geom_polygon(aes(x = long, y = lat, group = group), 
               color = "darkgray", fill = "lightgray") +
  coord_map(projection = "mollweide")
```

## Dataset `state`

```{r}
state <- map_data("state")
head(state, 10)
```

## Mapping `state`

```{r, fig.height = 3.5, fig.align = "center"}
ggplot(state) + 
  geom_polygon(aes(x = long, y = lat, group = group), 
               color = "darkgray", fill = "lightgray") +
  coord_map()
```

## Dataset `county`

```{r}
county <- map_data("county")
head(county, 10)
```

## Mapping `county`

```{r, fig.height = 3.5, fig.align = "center"}
ggplot(county) + 
  geom_polygon(aes(x = long, y = lat, group = group), 
               color = "darkgray", fill = "lightgray") +
  coord_map()
```

## Subsetting `state` for NC

```{r, fig.height = 3.5, fig.align = "center"}
ggplot(filter(state, region == "north carolina")) + 
  geom_polygon(aes(x = long, y = lat, group = group), 
               color = "darkgray", fill = "lightgray") +
  coord_map()
```

## Subsetting `county` for NC

```{r, fig.height = 3.5, fig.align = "center"}
ggplot(filter(county, region == "north carolina")) + 
  geom_polygon(aes(x = long, y = lat, group = group), 
               color = "darkgray", fill = "lightgray") +
  coord_map()
```

## Subsetting `world` for US

```{r}
world <- map_data("world")
str(world)
world_usa <- subset(map_data("world"), region == "USA")
```

## Mapping `world_usa`

```{r, fig.height = 3, fig.align = "center"}
ggplot(world_usa) + 
  geom_polygon(aes(x = long, y = lat, group = group), 
               color = "darkgray", fill = "lightgray") +
  coord_map()
```

## Adjusting Our X-Axis

```{r, fig.height = 3, fig.align = "center"}
ggplot(world_usa) + 
  geom_polygon(aes(x = long, y = lat, group = group), 
               color = "darkgray", fill = "lightgray")+
  coord_map(orientation = c(90, 180, 0))
```

## A Different Adjustment 

```{r, fig.height = 3, fig.align = "center"}
world_usa <- world_usa %>% 
  mutate(long_adj = ifelse(long > 0, long - 360, long))
ggplot(world_usa) + 
  geom_polygon(aes(x = long_adj, y = lat, group = group), 
               color = "darkgray", fill = "lightgray") +
  coord_map()
```

## The `mollweide` Projection

```{r, fig.height = 3.25, fig.width = 6, fig.align = "center"}
ggplot(world_usa) + 
  geom_polygon(aes(x = long_adj, y = lat, group = group), 
               color = "darkgray", fill = "lightgray") +
  coord_map(projection = "mollweide")
```

## The `ortho` Projection

```{r, fig.height = 3.25, fig.width = 6, fig.align = "center"}
ggplot(world_usa) + 
  geom_polygon(aes(x = long_adj, y = lat, group = group), 
               color = "darkgray", fill = "lightgray") +
  coord_map(projection = "ortho")
```

## Regions in `world_usa`

```{r}
str(world_usa)
```

***

```{r}
tail(world_usa %>%  count(subregion), 12)
```

## Mapping Hawaii

```{r, fig.height = 3.5, fig.align = "center"}
ggplot(filter(world_usa, subregion == "Hawaii")) + 
  geom_polygon(aes(x = long, y = lat, group = group), 
               color = "darkgray", fill = "lightgray") +
  coord_map()
```

## Ditch the Axes!

```{r}
# Theme Credit: http://eriqande.github.io/rep-res-web/lectures/
# making-maps-with-R.html (w/ the addition of panel.background)
# See also: https://ggplot2.tidyverse.org/reference/theme.html

ditch_the_axes <- theme(
  axis.text = element_blank(),
  axis.line = element_blank(),
  axis.ticks = element_blank(),
  panel.border = element_blank(),
  panel.grid = element_blank(),
  axis.title = element_blank(),
  panel.background = element_blank())
```

## Voila, a plot with no axes!

```{r, fig.height = 3, fig.align = "center"}
ggplot(filter(county, region == "north carolina")) + 
  geom_polygon(aes(x = long, y = lat, group = group), 
               color = "darkgray", fill = "lightgray") +
  coord_map() +
  ditch_the_axes
```

## A More Elaborate Version

```{r}
no_axes_alt <- 
  theme(panel.grid.major = element_line(colour = 'transparent'),
  axis.title.x = element_blank(), 
  axis.text.x = element_blank(),
  axis.ticks.x = element_blank(),
  axis.title.y = element_blank(), 
  axis.text.y = element_blank(),
  axis.ticks.y = element_blank(),
  panel.background = element_blank(),
  panel.border = element_blank(),
  panel.grid.minor = element_blank(),
  plot.background = element_blank())
```


## 
```{r, fig.height = 3.5, , fig.width = 6, fig.align = "center", include = FALSE}
ggplot(state) + 
  geom_polygon(aes(x = long, y = lat, group = group), 
               color = "darkgray", fill = "lightgray") +
  theme(legend.position = "none")
```

## References

* https://ggplot2.tidyverse.org/reference/coord_map.html
* http://eriqande.github.io/rep-res-web/lectures/making-maps-with-R.html
* https://socviz.co/maps.html
* https://github.com/UrbanInstitute/urbnmapr

* https://www.cdc.gov/nchs/data-visualization/county-teen-births/index.htm
* https://data.cdc.gov/NCHS/NCHS-Teen-Birth-Rates-for-Age-Group-15-19-in-the-U/3h58-x6cd
* https://data.cdc.gov/NCHS/NCHS-Teen-Birth-Rates-for-Age-Group-15-19-in-the-U/3h58-x6cd/data


