---
title: "Data Science 1<br>Presentation Example"
author: "Jill Thomley"
date: '`r format(Sys.time(), "%B %d, %Y @ %I:%M %p")`'
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Section 1: Dino Map!

## The Paleobiology Database

* non-government, non-profit, publicly-accessible resource for paleontological data

* organized / operated by an international, multi-disciplinary group of paleobiological researchers

* primary goal to provide global, collection-based occurrence and taxonomic data for organisms of all geological ages, as well data services for independent development of analytical tools, visualization software, and applications

* broader goal is to encourage and enable collaborations that address large-scale paleobiological questions

* You can filter and download records [HERE](https://paleobiodb.org/classic/displayDownloadGenerator) in many formats


## Plot of Suborder Therpoda

```{r, echo = FALSE, fig.width = 4}
library(tidyverse)
library(maps)
library(ggthemes)

dinosTheropodaUS <- read_csv("https://paleobiodb.org/data1.2/occs/list.csv?base_name=Theropoda&cc=US&state=!Hawaii,Alaska&show=class,coords,loc")

usa <- map_data("usa")
ggplot(usa) + 
  geom_polygon(aes(x = long, y = lat, group = group), color = "darkgray", fill = "lightgray") +
  coord_map(projection = "mollweide") +
  geom_point(dinosTheropodaUS, mapping = aes(x = lng, y = lat, color = class)) +
  scale_color_manual(values = c("magenta", "purple")) +
  guides(color = guide_legend("")) +
  labs(title = "Fossils of Dinosaurs in Suborder Therpoda Found in the US",
       subtitle = "aves have more bird-like hips, while saurischia have more lizard-like hips") +
  theme(
    axis.text = element_blank(),
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    panel.border = element_blank(),
    panel.grid = element_blank(),
    axis.title = element_blank(),
    legend.position = "bottom"
  )
```

Some text here!